---
title: "Peak Performance: A Fitness Tracker Data Exploration"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Loading in the Data

```{r}
setwd("C:/Users/plw40/OneDrive/Documents/Spring23_School/INFX512/Project")
physiological <- read.csv("phys_cycles.csv", header = TRUE, sep = ",")
sleeps <- read.csv("sleeps_1year.csv", header = TRUE, sep = ",")
workouts <- read.csv("workouts_1year.csv", header = TRUE, sep = ",")
journal <- read.csv("journal.csv", header = TRUE, sep = ",")
```

Structure of the Datasets

```{r}
str(physiological)
str(sleeps)
str(workouts)
str(journal)
```

As you can see by the str() structure of the dataframes, I will need to convert the date and time variables to a date-time data type. I will also need to convert the type for the "color" variable from character to a factor with labels.

```{r}
install.packages("lubridate")
library(lubridate)
##converting character to datetime using lubridate
physiological$cstart <- mdy_hm(physiological$cstart)
physiological$cend <- mdy_hm(physiological$cend)
physiological$sleepstart <- mdy_hm(physiological$sleepstart)
physiological$wake <- mdy_hm(physiological$wake)
sleeps$cstart <- mdy_hm(sleeps$cstart)
sleeps$cend <- mdy_hm(sleeps$cend)
sleeps$sleeponset <- mdy_hm(sleeps$sleeponset)
sleeps$wakeonset <- mdy_hm(sleeps$wakeonset)
workouts$cstart <- mdy_hm(workouts$cstart)
workouts$cend <- mdy_hm(workouts$cend)
workouts$workoutstart <- mdy_hm(workouts$workoutstart)
workouts$workoutend <- mdy_hm(workouts$workoutend)
journal$cstart <- mdy_hm(journal$cstart)
journal$cend <- mdy_hm(journal$cend)

##converting physiological recovery "color" to factor
physiological$color <- as.factor(physiological$color)
str(physiological$color)
```

Now that the data is correctly categorized, I want to take a quick look at the summary statistics for each of the datasets.

```{r}
library(tidyverse)
summary(physiological)
summary(sleeps)
summary(workouts)
summary(journal)
```
Let's take a quick look at all of the physiological based on recovery grouping.

```{r}
group <- NA
group[physiological$recovery <= 33] <- 1
group[physiological$recovery >= 34 & physiological$recovery <= 66] <- 2
group[physiological$recovery >= 67] <- 3

pairs(physiological[ ,c(3,5:10,13,14)],
      col= c("red","yellow","green")[group],
      pch= c(15, 16, 17)[group],
      main = "Plot of Variables by Recovery Score Grouping")
```
Based on the scatterplots, I can see that when RHR (resting heart rate) is high, I seem to have low/red recovery days. When RHR is low, I have more high/green recovery days. HRV (heart rate variability), offers very distinct groupings. High HRV is associated with high/green recovery, mid-range HRV is associated with yellow/medium recovery, and low HRV is associated with low/red recovery. Daystrain, Calories, and MaxHR have positive linearity. Respiratory rate x HRV offers the most distinct groupings. I'd say based on a quick glance of this high-level overview that RHR, HRV, and Respiratory Rate will likely be good prediction variables when we get to predictive modeling. 
---
Let's get a closer look at some of the variables in direct relation to Recovery Score. 

```{r}
library(ggplot2)
recovery.rhr <- ggplot(data=physiological) +
  geom_point(mapping=aes(rhr, recovery, color = color))+
  scale_color_manual(name = "Recovery Score", values= c("red"="red",
                               "yellow"="yellow",
                               "green"="green")) +
  labs(title="Recovery and RHR")
recovery.rhr
```
When taking a closer look at Recovery and RHR we can see that a RHR of >= 65 results in a low/red recovery score.There appears to be a negative linear relationship between the two variables. 

```{r}
recovery.hrv <- ggplot(data=physiological) +
  geom_point(mapping=aes(hrv, recovery, color = color))+
  scale_color_manual(name = "Recovery Score", values= c("red"="red",
                               "yellow"="yellow",
                               "green"="green")) +
  labs(title="Recovery and HRV")
recovery.hrv
```
HRV and Recovery appear to have an obvious positive linear relationship. As HRV increases, so does Recovery.
```{r}
cor(physiological$recovery, physiological$hrv)
```
```{r}
recovery.rhr <- ggplot(data=physiological) +
  geom_point(mapping=aes(rhr, recovery, color = color))+
  scale_color_manual(name = "Recovery Score", values= c("red"="red",
                               "yellow"="yellow",
                               "green"="green")) +
  labs(title="Recovery and RHR")
recovery.rhr
```
```{r}
cor(physiological$recovery, physiological$rhr)
```



```{r}
recovery.sleep <- ggplot(data=physiological) +
  geom_point(mapping=aes(sleepscore, recovery, color = color))+
  scale_color_manual(name = "Recovery Score", values= c("red"="red",
                               "yellow"="yellow",
                               "green"="green")) +
  labs(title="Recovery and Sleep Score")
recovery.sleep
```
There is no linear relationship between Sleep and Recovery. However, these two variables will need further investigation to understand how they relate to one another. 
```{r}
cor(physiological$sleepscore, physiological$recovery)
```


```{r}
recovery.resprate <- ggplot(data=physiological) +
  geom_point(mapping=aes(resprate, recovery, color = color))+
  scale_color_manual(name = "Recovery Score", values= c("red"="red",
                               "yellow"="yellow",
                               "green"="green")) +
  labs(title="Recovery and Respiratory Rate")
recovery.resprate
```
```{r}
cor(physiological$recovery, physiological$resprate)
```

I'd like to see my recovery scores displayed by their color indicator over the span of this year of data. I will first make a dataframe with just the relevant information.
```{r}
##create new dateframe for daily recovery
dailyrecovery = data.frame(
  date=as.Date(physiological$cstart),
  recovery= as.integer(physiological$recovery),
  color= as.character(physiological$color))

library(dplyr)
##creating a dummy variable representing the 3 levels of recovery 0=red recovery days, 50=yellow recovery days, and 100=green recovery days
dailyrecovery <- dailyrecovery %>% 
  mutate(recovery_level = case_when(
    recovery <= 33~0, 
    recovery >= 34 & recovery <=66 ~50, 
    recovery >= 67 ~ 100))
head(dailyrecovery) 
```
```{r}
library(calendR)
recovery_scores <- calendR(start_date = "2020-04-01",
                           end_date = "2021-03-31",
                           start= "M",
                           title= "My Daily Recovery",
                           special.days = dailyrecovery$recovery_level,
                           special.col= "green",
                           gradient=TRUE,
                           low.col="red",
                           weeknames = c("S","M","T","W","T","F","S"),
                           day.size = 1.8,
                           font.style = "bold"
                           )
plot(recovery_scores)
```
```{r}
table(physiological$color)
```

In order to create this calendar, I first created a dataframe containing the information about my recovery by the day. I then had to create a dummy column with 3 distinct values to work in correlation with the color of the recovery that day. Since recovery is coded as red, yellow, and green I used the values 0,50,100 in my dummy variable to correlate to red=low where recovery is <=33, yellow=moderate where recovery is between 34 and 66, and green=high where recovery is >=67. I then plot this on a calendar using the calendR package. The result is a calendar representing my recovery score color for each day over the year of data. Here, it is easy to see that I had very few low recovery (red) days in this year. We will look more into the contributing factors to these days later on. 

---
```{r}
pairs(sleeps[,5:12],
      col="dodgerblue3",
      pch= 20,
      main = "Sleep Data Plots")
```
See positive linear relationships among Sleep Score and time: spent asleep, in bed, light sleep, deep sleep, and REM sleep. 

```{r}
plot(sleeps$cstart,
     sleeps$lightsleep,
     type= "l",
     col= "lightblue",
     xlab= c("Months"),
     ylab= c("Minutes"),
     main= "Sleep Cycles")
lines(sleeps$cstart,
      sleeps$deepsleep,
      type= "l",
      col= "midnightblue")
lines(sleeps$cstart,
      sleeps$rem,
      type="l",
      col="dodgerblue")
legend("topright",
       c("light sleep","deep sleep", "REM sleep"),
       lty = 1,
       col= c("lightblue", "midnightblue", "dodgerblue"))
```
Based on this it is easy to see that I spend the least amount of time in deep sleep and on average more time in light sleep than any other sleep cycle. 

```{r}
avg.light <- mean(sleeps$lightsleep)
avg.deep <- mean(sleeps$deepsleep)
avg.rem <- mean(sleeps$rem)
avg.awake <- mean(sleeps$awake)
avg.total <- avg.awake+avg.deep+avg.light+avg.rem

cycles <- data.frame(
  category=c("light", "deep", "REM", "awake"),
  time=c(avg.light, avg.deep, avg.rem, avg.awake))

cycles$fraction <- cycles$time / avg.total
cycles$fraction <- round(cycles$fraction, digits = 2)

pie(cycles$fraction, main="Sleep Cycles", 
    labels=c("light 40%", "deep 19%", "REM 33%", "awake 8%"),
    col=c("lightblue", "midnightblue", "dodgerblue", "red"))
```
```{r}
ggplot(sleeps) +
  geom_bar(aes(x=cstart, y=asleep), stat="identity", fill="lightblue", color="lightblue") +
  geom_line(aes(x=cstart, y=sleepneed), stat="identity", color="orange") +
              labs(title="Sleep Performance", x="Months", y="Minutes")
  
```
```{r}
ggplot(sleeps) +
  geom_bar(aes(x=cstart, y=inbed), stat="identity", fill="orange") +
  geom_bar(aes(x=cstart, y=asleep), stat="identity", fill="lightblue") +
  labs(title="Sleep Efficiency", x="Months", y="Minutes")
```


```{r}
ggplot(workouts, aes(x=activity)) + 
  geom_bar(color="darkorange", fill="darkorange")+
  labs(x= "Activity", y= "Count", title= "Types of Workouts")+
  theme(axis.text.x = element_text(angle = 90))
  
```
```{r}
avg.strain <- aggregate(activitystrain~activity, data=workouts, FUN=mean)
avg.strain$activitystrain <- round(avg.strain$activitystrain, digits= 1)
ggplot(avg.strain, aes(x=activity, y=activitystrain)) +
  geom_bar(stat="identity", fill="lightblue") +
  labs(title="Avg. Activity Strain by Activity Type", x="Activity Type", y="Average Strain") +
  geom_text(aes(label=activitystrain), vjust= 1) +
  theme(axis.text.x = element_text(angle = 90))
```
```{r}
hrz.activity <- read.csv("hrzactivity.csv", header = TRUE, sep = ",")
```

```{r}
ggplot(hrz.activity, aes(fill=group, y=value, x=activity)) + 
  geom_bar(position="fill", stat="identity") +
  theme(axis.text.x = element_text(angle = 90),panel.background = element_blank()) +
  labs(title="HRZ by Activity") +
  scale_fill_manual(values=c('grey90', 'grey70', 'steelblue', 'seagreen3','sandybrown','orangered'))
```


Recovery v Day Strain v Sleep v HRV
```{r}
plot(physiological$cstart,
     physiological$recovery,
     type= "l",
     col= "forestgreen",
     xlab= c("Months"),
     ylab= c("Value"),
     main= "Key Metrics Over Time")
lines(physiological$cstart,
      physiological$sleepscore,
      type= "l",
      col= "midnightblue")
lines(physiological$cstart,
      physiological$hrv,
      type="l",
      col="mediumpurple")
legend("topright",
       c("Recovery Score","Sleep Score", "HRV"),
       lty = 1,
       col= c("forestgreen", "midnightblue", "mediumpurple"))
```
```{r}
summary(physiological$hrv)
```
```{r}
summary(physiological$rhr)
```

```{r}
summary(physiological$sleepscore)
```
```{r}
summary(physiological$resprate)
```
```{r}
summary(sleeps$sleepneed)
```
```{r}
summary(sleeps$sleepdebt)
```
```{r}
summary(sleeps$efficiency)
```
```{r}
summary(sleeps$inbed)
```


```{r}
summary(sleeps$asleep)
```
```{r}
summary(workouts$duration)
```
```{r}
##creating a red day data frame
attach(physiological)
red.days <- physiological[which(recovery <= 33),]
detach(physiological)
head(red.days)
```
summary stats on red days
```{r}
summary(red.days[ ,c(3:10, 13:23)])
```
red workout days


Red Day Journals
```{r}
red.journal <- merge(red.dates, journal, by = "cstart")
str(red.journal)
```


High Green dataframe

```{r}
##creating high green data frame
attach(physiological)
high.green <- physiological[which(recovery >= 90),]
detach(physiological)
high.dates <- as.data.frame(high.green[,1])
colnames(high.dates)[1] = "cstart"
high.dates
```
```{r}
summary(high.green[ ,c(3:10, 13:23)])
```
high green workouts
```{r}

head(high.workouts)
```

High Green Journal Entries
```{r}
high.journal <- merge(high.dates, journal, by= "cstart")
str(high.journal)
```
overall phys stats
```{r}
summary(physiological[ ,c(3:10, 13:23)])
```
reading in physiological comparison .csv
```{r}
##average physiological stat comparison chart for visualization
physavg.comparison <- read.csv("physiologicalcomparison.csv", header = TRUE, sep = ",")
head(physavg.comparison)
```
Visualizing differences red, high, overall
```{r}
physavg.1 <- physavg.comparison[c(1,2,6,7,12,14,17,18,19,23,24,29,31,34,35,36,40,41,46,48,51), ]
physavg.2 <- physavg.comparison[c(3,8,16,20,25,33,37,42,50), ]
physavg.3 <- physavg.comparison[c(5,11,13,22,28,30,39,45,47), ]
physavg.4 <- physavg.comparison[c(9,10,15,26,27,32,43,44,49), ]
physavg.5 <- physavg.comparison[c(4,21,38), ]
```


```{r}
ggplot(physavg.1, aes(fill=group, y=value, x=group)) + 
  geom_bar(position="dodge", stat="identity") +
  theme(panel.background = element_blank()) +
  labs(title="Average Physiological Metrics Comparison 1", x="metric") +
  scale_fill_manual(values=c('seagreen3', 'steelblue', 'orangered')) +
  facet_wrap(~stat)
```
group 2
```{r}
ggplot(physavg.2, aes(fill=group, y=value, x=stat)) + 
  geom_bar(position="dodge", stat="identity") +
  theme(panel.background = element_blank()) +
  labs(title="Average Physiological Metrics Comparison 2", x="metric") +
  scale_fill_manual(values=c('seagreen3', 'steelblue', 'orangered'))
```
group 3
```{r}
ggplot(physavg.3, aes(fill=group, y=value, x=stat)) + 
  geom_bar(position="dodge", stat="identity") +
  theme(panel.background = element_blank()) +
  labs(title="Average Physiological Metrics Comparison 3", x="metric") +
  scale_fill_manual(values=c('seagreen3', 'steelblue', 'orangered'))
```
comparison 4
```{r}
ggplot(physavg.4, aes(fill=group, y=value, x=stat)) + 
  geom_bar(position="dodge", stat="identity") +
  theme(panel.background = element_blank()) +
  labs(title="Average Physiological Metrics Comparison 4", x="metric") +
  scale_fill_manual(values=c('seagreen3', 'steelblue', 'orangered'))
```
comparison 5
```{r}
ggplot(physavg.5, aes(fill=group, y=value, x=stat)) + 
  geom_bar(position="dodge", stat="identity") +
  theme(panel.background = element_blank()) +
  labs(title="Average Physiological Metrics Comparison 5", x="metric") +
  scale_fill_manual(values=c('seagreen3', 'steelblue', 'orangered'))
```


```{r}
phys.numeric <- physiological[ ,c(3, 5:10, 13:23)]
head(phys.numeric)
```


```{r}
library(ggcorrplot)
phys.cor <- cor(phys.numeric) 
cor.plot <- ggcorrplot(phys.cor, hc.order = TRUE, type="lower", lab=TRUE) +
  labs(title="Correlation of Physiological Variables")
cor.plot
```

exploring workouts for high and low days
```{r}
high.workouts <- merge(high.dates, workouts, by = "cstart")
high.workouts <- high.workouts[ , -2:-4] ##removing redundant dates
## adding recovery score to the df
df.high <- merge(high.workouts, high.green[ ,c("cstart","recovery")], by="cstart", all.x=TRUE)
df.high <- df.high[ ,c("cstart", "recovery", "duration", "activity", "activitystrain", "calories", "maxhr", "avghr", "hrz1", "hrz2", "hrz3", "hrz4", "hrz5")]
high.workouts <- df.high
head(high.workouts)
```
```{r}
high.workouts$cstart <- as.character(high.workouts$cstart)
ggplot(high.workouts, aes(x=cstart)) +
  geom_bar(fill="seagreen3")+
  theme(axis.text.x = element_text(angle = 90),panel.background = element_blank()) +
  labs(title="# of Activities per Day", x="date")
```
```{r}
attach(high.workouts)
strain.day <- aggregate(activitystrain~cstart, data=high.workouts, FUN=sum)
ggplot(strain.day, aes(x=cstart, y=activitystrain)) +
  geom_bar(stat="identity", fill="orangered")+
  theme(axis.text.x = element_text(angle = 90),panel.background = element_blank()) +
  labs(title="Total Activity Strain by Day", x="date")
```


exploring workout data for red days

```{r}
red.dates <- as.data.frame(red.days[,1])
colnames(red.dates)[1] ="cstart"
red.workouts <- merge(red.dates, workouts, by="cstart")
red.workouts <- red.workouts[,-2:-4] ##removing redundant dates
red.workouts <- merge(red.workouts, red.days[ ,c("cstart", "recovery")], by="cstart", all.x=TRUE)
red.workouts <- red.workouts[ ,c("cstart", "recovery", "duration", "activity", "activitystrain", "calories", "maxhr", "avghr", "hrz1", "hrz2", "hrz3", "hrz4", "hrz5")]
head(red.workouts)
```
```
```{r}
red.journal <- red.journal[,-2]
high.journal <- high.journal[,-2]
```
---
```{r}
ggplot(data= red.journal, aes(x=question, y=answeredyes, fill=answeredyes)) +
  geom_bar(stat="identity") +
  coord_flip() +
  labs(title="Journal Entries for Red Days", fill="Response") +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
  scale_fill_manual(values=c("orangered","steelblue"))
```
---
```{r}
ggplot(data= high.journal, aes(x=question, y=answeredyes, fill=answeredyes)) +
  geom_bar(stat="identity") +
  coord_flip() +
  labs(title="Journal Entries for High Green Days", fill="Response") +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
  scale_fill_manual(values=c("orangered","steelblue"))
```

##predictive modeling: simple linear regression
```{r}
## Splitting the data
set.seed(1)
sample.set <- sample(x=nrow(phys.numeric), size=0.7*nrow(phys.numeric))
slm.train <- phys.numeric[sample.set, ]
slm.test <- phys.numeric[-sample.set, ]

## Running simple linear regression
simplelm <- lm(recovery~hrv, data=slm.train)
summary(simplelm)
```

